---
output: github_document
---

![Build Status](https://travis-ci.org/mastoffel/partR2.svg?branch=master)
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# partR2

The goal of partR2 is to implement methods for LMMs and GLMMs to estimate the R2
along with it's confidence intervals as well as partition the variation explained unique to and common
among certain predictors, which is useful in general but particularly
in the case of multicollinearity. All confidence intervals are estimated using
parametric bootstrapping. 

partR2 is in an early phase of development.

## Installation

You can install partR2 from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mastoffel/partR2")
```

## Example

```{r example, results = TRUE, warning=FALSE}
library(partR2)
library(lme4)

?partR2

# load data
data(BeetlesBody)
# fit lme4 model
mod <- lmer(BodyL ~ Sex + Treatment + Habitat + (1|Container) + (1|Population),
            data = BeetlesBody)

# partition R2
R2 <- partR2(mod, partvars = c("Sex", "Treatment", "Habitat"), 
                   R2_type = "marginal", nboot = 50, CI = 0.95)
R2
```

And to plot the results:
```{r plot, results = TRUE, warning=FALSE}
forestplot(R2, type = "R2")
```
