---
output: github_document
---
<!-- badges: start -->
![Build Status](https://travis-ci.org/mastoffel/partR2.svg?branch=master)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Codecov test coverage](https://codecov.io/gh/mastoffel/partR2/branch/master/graph/badge.svg)](https://codecov.io/gh/mastoffel/partR2?branch=master)
<!-- badges: end -->
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r echo=FALSE, results="hide", message=FALSE}
#library("txtplot")
library("badger")
```

# partR2

The goal of partR2 is to estimate R^2^ in GLMMs (sensu Nakagawa & Schielzeth 2013) and to partition the R^2^ into the variance explained by the predictors.

The package takes a fitted lme4 model as input and gives you:

* R^2^ (marginal or conditional)
* Semi-partial R^2^, the variance explained uniquely by each predictor and combinations of predictors
* Inclusive R^2^, the variance explained by a predictor independent of all
other predictors
* Structure coefficients, the correlation between a predictor and the fitted response
* Beta weights, standardised model estimates

All estimates can be combined with parametric bootstrapping to get confidence intervals.

partR2 is in an early phase of development and despite a lot of testing might still contain bugs, so 
be careful and thoughtful when using it. 

## Installation

You can install partR2 from github with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("mastoffel/partR2", build_vignettes = TRUE, dependencies = TRUE) 
# check vignette
browseVignettes("partR2")
```

## Example

```{r example, results = TRUE, warning=FALSE, message=FALSE}
library(partR2)
library(lme4)

?`partR2-package`

# load data
data(biomass)
# fit lme4 model
mod <- lmer(Biomass ~  Year + Temperature + SpeciesDiversity + (1|Population),
            data = biomass)
# R2s and partial R2s
(R2 <- partR2(mod,  partvars = c("SpeciesDiversity", "Temperature", "Year"),
              R2_type = "marginal", nboot = 100, CI = 0.95))
```

And to plot the results:
```{r plot, results = TRUE, warning=FALSE, fig.width=7, fig.height=4}
forestplot(R2, type = "R2", line_size = 0.7, text_size = 14, point_size = 3)
```


### Citation

The package is still in development, but will hopefully be uploaded to CRAN soon.

When using it, please cite our [preprint](https://www.biorxiv.org/content/10.1101/2020.07.26.221168v1.abstract) for now:

Stoffel, MA, Nakagawa, S, & Schielzeth, H (2020). partR2: Partitioning R2 in generalized linear mixed models. bioRxiv. 
`r badge_doi("10.1101/2020.07.26.221168", "green")`
